---
title: "Tables of abundance and basal area"
subtitle: BCI, Sherman and Cocoli
author: "Mauro Lepore"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    theme: united
    keep_md: true
---

```{r setup, include=FALSE}
# This chunk named setup will run before any other code (https://goo.gl/BeM2Uu)
set.seed(1014)
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  # Figures (http://r4ds.had.co.nz/)
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,
  fig.show = "hold"
)
```

Setup.

```{r packages} 
library(tidyverse)
library(fgeo)
```

To create tables of abundance and basal area first pick the data you want. Your code should look something like this:

```R
pick1 <- pick_plotname(VIEWFULLTABLE, "PLOTNAME")
pick2 <- pick_dbh_min(pick1, MINIMUM_DBH)
```

Then you can create the tables with:

```R
abundance <- abundance_byyr(pick2)
basal_area <- basal_area_byyr(pick2)
```

To standardize the table of basal area use something like this:

```R
years <- setdiff(names(basal_area), c("species", "Family"))
in_sq_m <- convert_unit_at(basal_area, .at = years, from = "mm2", to = "m2")
basal_area_in_he <- standardize_at(in_sq_m, .at = years, denominator = denominator_HECTARES)
```

Next, I do this for three plots: Sherman, Cocoli and BCI. To avoid duplication I first write some disposable utility-funcitons.

```{r utils} 
# Pick specific data from a ViewFullTable.
pick_vft <- function(vft, plot_nm) {
  pick <- pick_plotname(vft, plot_nm)
  pick_dbh_min(pick, 10)
}

# Standardize basal area by total plot-hectares.
standardize_ba <- function(ba, denominator) {
  years <- setdiff(names(ba), c("species", "Family"))
  in_he <- convert_unit_at(ba, .at = years, from = "mm2", to = "m2")
  standardize_at(in_he, years, denominator)
}

path_iss59 <- function(path) {
  here::here("inst/issues/59_abund_tables", path)
}

write_tables <- function(vft, plot_nm, denominator) {
  pick <- pick_vft(vft, plot_nm)
  
  abun <- abundance_byyr(pick)
  readr::write_csv(abun, path_iss59(paste0("tbl/", plot_nm, "_abundance.csv")))
  
  ba <- standardize_ba(basal_area_byyr(pick), denominator)
  readr::write_csv(ba, path_iss59(paste0("tbl/", plot_nm, "_basal_area.csv")))
}
```

Now I'm ready to read the data and crete the tables -- which I save locally:

* Read ViewFullTables.

```{r data} 
# BCI
path_bci <- here::here("inst/issues/59_abund_tables/vft_bci.csv")
vft_bci <- readr::read_csv(path_bci)

# Sherman and Cocoli
path_sc <- here::here("inst/issues/59_abund_tables/vft_sc.csv")
vft_sc <- readr::read_csv(path_sc)
```

* Write tables of abundance and basal area.

```{r}
write_tables(vft_sc, "sherman", denominator = 5.96)
write_tables(vft_sc, "cocoli", denominator = 4)
write_tables(vft_bci, "bci", denominator = 50)
```

Les't visualize the output.

```{r, message=FALSE}
fgeo.tool::csv_to_df_lst(here::here("inst/issues/59_abund_tables/tbl"))
```


