---
title: "Mapping species distribution"
author: "Mauro Lepore"
date: "2017-11-01"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{Mapping species distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
# hadley's settings
set.seed(1014)
options(digits = 3)

knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)

library(ggplot2)
update_geom_defaults("text", list(size = 14.5))
```

This vignette show you how to map species distribution with `map_sp()` and `map_sp_pdf()` from the __forestr__ package. Let's load __forestr__:

```{r}
GITHUB_PAT <- "your token"
# install_github("forestgeo/forestr", auth_token = GITHUB_PAT)
library(forestr)

# Also using the package dplyr for easier data manipulation;
# it is the most downloaded R package (https://goo.gl/bXBmkS).
library(dplyr)
# Print only a few rows of tibbles (modern dataframes) to save space
options(dplyr.print_min = 6, dplyr.print_max = 6)
```

__forestr__ has some data that you will use for examples:

```{r}
example_data <- dplyr::as_tibble(forestr::bci12t7mini)

# Let's find the top-3 most abundant species
sp_n <- count(example_data, sp)
arranged <- arrange(sp_n, desc(n))
sp_top3 <- head(arranged, 3)$sp
sp_top3

# For a this example, let's filter the data set to keep only the top-3 species
census <- filter(example_data, sp  %in% sp_top3)
census
```

The first section shows you how to explore your data quickly and yet get good quality maps with the available defaults. The second section shows you how to customize you maps; It is organized to show first the changes you may need for exploring your data, and second the changes you may need to precisely communicate your results.

# Exploring Data Quickly: Using Defaults

## Species Distribution

Although the output of `map_sp()` and `map_sp_pdf()` is printed to different devices -- i.e, the screen or a .pdf file --  those functions are almost identical. They are documented jointly and share the same arguments. 

```
?map_sp
# Or
?map_sp_pdf
```

Notice that only the first two arguments of these functions are strictly necessary.

```{r}
# Print to screen
map_sp(census, c("hybapr", "faraoc"))

# Save in a .pdf (defaults to working directory)
map_sp_pdf(census, c("hybapr", "faraoc"))
```

Keep in mind that your census data must have specific variable-names:

```{r, error=TRUE}
# Show behaviour with wrong name
with_wrong_name <- rename(census, SP = sp)

# This will fail because the data set has name `SP`, not `sp`
map_sp(with_wrong_name, c("hybapr", "faraoc"))
```

## Elevation

You may have elevation data from your site.

```{r}
my_elev <- bciex::bci_elevation
my_elev
```

Note that the elevation data set must have very specific variable names; or it will fail:

```{r, error=TRUE}
# This will fail because the names of `my_elev` are not correct
map_sp(census, c("hybapr", "faraoc"), elevation = my_elev)
```

If necessary, rename your variables:

```{r}
# Renaming data and trying again
elev <- rename(my_elev, gx = x, gy = y)
map_sp(census, c("hybapr", "faraoc"), elevation = elev)
```

# Customizing Your Maps

Eventually you may want to customize the output of `map_sp()` and `map_sp_pdf()`. When you are exploring your data you may want to quickly change a few things such as the some property of the points, the general theme of your map; or the colour of the elevation lines. And eventually you may want to fine tune your map for publication. This section helps you to do all of that.

## Most Common Changes

### Changing Ponint Properties

Recall that both `map_sp()` and `map_sp_pdf()` have the same arguments; the argument `...`, in particular, allows you to pass arguments to `ggplot2::geom_point()`. Read `?ggplot2::geom_point()` to learn all your options; here you can see some common ones:

```{r}
map_sp(census, "hybapr",
  # Arguments passed to ggplot2::geom_point()
  size = 4, shape = 22, fill = "green", colour = "black", stroke = 2)
```

### Changing the Theme

You can change the overall look of your map with a pre-set ggplot theme. Learn about the available themes by reading `?ggplot2::theme_bw()`.

```{r}
map_sp(census, "hybapr", theme = ggplot2::theme_classic())
```

### Changing the Elevation Lines

You can also customize the elevation lines. This example shows all the available options:
* Change line size;
* set colours to represent low and high elevation;
* set the number of intervals in which the elevation variable is to be cut (note that the cut results in ugly numbers, with many decimals).

```{r}
map_sp(census, "hybapr", 
  elevation = elev, line_size = 1, low = "red", high = "blue", bins = 2)
```

## Less Common Changes

Here are some changes that you may need less often but are still good to keep in mind.

### Change File Name

When printing to .pdf you can provide a custom file name (and path to a directory); but note the file name must end in _.pdf_ or it will be replaced by a default name:

```{r, error=TRUE}
spp <- c("hybapr", "faraoc")
map_sp_pdf(census, spp, elevation = elev, file = "bad-name1")
map_sp_pdf(census, spp, elevation = elev, file = "bad-name2.png")
map_sp_pdf(census, spp, elevation = elev, file = "good-name.pdf")
```

### Change x and y Limits

Occationally you may need to change the map limits. The map limits default to the range 0-max, where max is the maximum value of the variable `gx` or `gy` in the census data. 

In the rare case where no stem occurrs at the maximum limits of your plot, the default would result in a map that spans less than the defined limits of your data.

```{r}
# Fake case: The plot maximum-limits are gx = 1000 and gy = 500; but all stems
# in the census occurr at lower gx and gy limits
rare_case <- filter(census, gx < 800, gy < 200)
map_sp(rare_case, unique(rare_case$sp))
```

You can overwrite the default limits with the arguments `xlim` and `ylim`:

```{r}
map_sp(rare_case, unique(rare_case$sp), xlim = c(0, 1000), ylim = c(0, 500))
```

# Fine Tunning

This last section shows you how to fine tune your plots. For this, you will need a couple of functions from packages other than __forestr__. Because you this kind of changes are rare you may forget how to do them, but you can come back to notes whenever you

## Multiple Plots in One Page

To plot multiple maps on one page you can use `gridExtra::maparrange()`.

```{r}
library(gridExtra)
maps <- map_sp(census, sp_top3)
multipaged <- marrangeGrob(map_sp(census, sp_top3), nrow = 1, ncol = 2)
multipaged
```

Now you can save your multi-paged maps to a .pdf

```{r}
# Saving to .pdf: Option 1
ggplot2::ggsave("multi-paged.pdf", multipaged)

# Saving to .pdf: Option 2
pdf()
multipaged
dev.off()
```

You can also print to other devices. If your maps span multiple pages, .png() will create one file per page.

```{r}
# Saving to .png; this will create multiple files.
png()
multipaged
dev.off()
```


## Fine Tunning the Theme

You may have noticed that the default value of the argument `theme` is a function from the __ggplot__ package. Before you used a pre-made theme; but you can create your own theme with `ggplot2::theme()` and then pass it to `map_sp()` or `map_sp_pdf()`. 


```{r}
# Showing only a few options; see all the options with ?ggplot2::theme()
mytheme <- ggplot2::theme(
  panel.background = element_rect(fill = "lightgreen"),
  panel.grid.minor = element_line(linetype = "dashed"),
  strip.background = element_rect(fill = "black"),
  strip.text = element_text(colour = "white"),
  text = element_text(size = 25, face = "bold")
)
map_sp(census, "hybapr", theme = mytheme)
```

__ggplot2__ is the engine of `map_sp()` and `map_sp_pdf()`, but these functions are built so you can use them without konwing anything about __ggplot2__. Yet you may want to know more about it. __ggplot2__ is not an arbitrary collection of functions that you need to memorize; instead it is one implementation of [The Grammar of Graphics](https://goo.gl/E514ar). You can learn __ggplot2__ from [this website](https://goo.gl/tVvLQZ); this specialized, paid book: [ggplot2: Elegant Graphics for Data Analysis](https://goo.gl/R6wJk9); or this general, free book [R for Data Science](http://r4ds.had.co.nz/). Learning always has some overhead, but the learning curve of __ggplot2__ is not steep. It pays off.
