---
title: "Count saplings"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(fgeo)
```

Data.

```{r}
census <- data.frame(
    stringsAsFactors = FALSE,
    dbh = c(20, 120, 22, 99, NA),
    sp = c("sp1", "sp1", "sp2", "sp2", "sp2"),
    treeID = c("1", "1", "2", "2", "2"),
    stemID = c("1.1", "1.2", "2.1", "2.2", "2.3")
)
census
```

> For example, if I want to see how many saplings there are per species, the
dbh range would be >=10mm and <100mm. A tree with multiple stems 120 and 20 will be counted, is this correct?

Counting saplings by species.

```{r}
# This preserves missing `dbh` (unless `na.rm = FALSE`)
saplings <- census %>% 
  pick_dbh_min(10) %>% 
  pick_dbh_under(100)
saplings

# Count unique instances of treeID by species
saplings %>% 
  group_by(sp) %>% 
  abundance_tree()
```

Bonus:

* Other ways of filtering desired dbh.

```{r}
saplings <- filter(census, dbh >= 10, dbh < 100)
# Same
saplings <- census %>% 
  filter(dbh >= 10, dbh < 100)
# Same
saplings <- census %>% 
  filter(dbh >= 10) %>% 
  filter(dbh < 100)
saplings
```

* Other counts.

```{r}
# Count unique instances of treeID
abundance_tree(saplings)

# Count unique instances of stemID by species
saplings %>% 
  group_by(sp) %>% 
  abundance_stem()
```

> Then I want to count trees per species, i.e. >=100mm, this tree would also be counted by your current function, is this correct?

Counting trees by species.

```{r}
# This preserves missing `dbh` (unless `na.rm = FALSE`)
trees <- census %>% 
  pick_dbh_min(100)
trees

# Count unique instances of treeID by species
trees %>% 
  group_by(sp) %>% 
  abundance_tree()
```

> If so, this will be wrong, since it will be counted twice. A tree with stems 120 and 20 is a tree, not a sapling, and should only be counted as a tree.

You are right. Thanks for your example and expectation. Now I see the problem. The good solution seems to be a specialized function that collapses a table of potentially multi-stem trees to a table of one stem per tree, where the stem that remains is that of maximum `dbh`. The function may look someting like this:

```{r}
collapse_dbh_max <- function(x) {
  x %>% 
    group_by(treeID) %>% 
    arrange(desc(dbh)) %>% 
    filter(row_number() == 1) %>% 
    ungroup()
}
```

Now, we start by collapsing to maximun dbh and then we repeat the process/

```{r}
saplings <- census %>% 
  collapse_dbh_max() %>% 
  filter(dbh >= 10) %>% 
  filter(dbh < 100)
saplings

saplings %>% 
  group_by(sp) %>% 
  abundance_tree()
```

```{r}
trees <- census %>% 
  collapse_dbh_max() %>% 
  filter(dbh >= 100)
trees

trees %>% 
  group_by(sp) %>% 
  abundance_tree()
```

Is this the expected result?


TODO: 

* fix warnings
* wrap into `count_trees()` and `count_saplings()`, by ... or `count_between(.data, ..., lower = 0, upper = Inf)`.
* move this article out of tutorial into issues.
