---
title: "Basal area: Comparing output from Suzanne and mine"
subtitle: "Check why species with buttressess have basal area higher than expected"
output: github_document
---

Suzanne wrote:

> Most of the numbers coincide. However, for some species (i.e. buttressed trees), they do not: Ceiba pentandra, Alseis blackiana. Dipteryx oleifera, etc. The difference in census 3 is quite high. Can you please check to see why you got such a high number for this census? The numbers after this census are also quite high.

> Remember that you have to make a table with one record per census per stemid. The stems from buttressed trees may be measured more than once in a census but at different heights. So you have to choose the record where the HOM is hightest for any stemid. I think that you just summed all the records without taking this into consideration.

Thanks Suzanne. I had a bug here: When I changed the code to sum all stems in each tree (within the given DBH range) I unintentionally disabled the code that picks the (single) StemID with largest HOM. Now that's fixed and added a formal test so this doesn't happen again. Now the basal area of buttressed species should be lower than before.

```{r setup, include=FALSE}
# This chunk named setup will run before any other code (https://goo.gl/BeM2Uu)
set.seed(1014)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  # Figures (http://r4ds.had.co.nz/)
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,
  fig.show = "hold"
)
```

```{r packages, eval=TRUE, message=FALSE, warning=FALSE} 
library(tidyverse)
library(fgeo)
```

Outputs by me and Suzanne

```{r data, message=FALSE}
path_mao <- here::here(
  "inst/issues/59_abund_tables/out/bci_sapling_tree_basal_area.csv"
)
mao <- readr::read_csv(path_mao)

path_lao <- here::here("inst/issues/59_abund_tables/lao/BasalArea.xlsx")
lao <- readxl::read_excel(path_lao)
lao <- unite(lao, "species", Genus, SpeciesName, sep = " ")
```

Some species that have buttresses.

```{r}
spp <- c("Ceiba pentandra", "Alseis blackiana", "Dipteryx oleifera")
spp
```

Helper funciton to reduce duplication.

```{r}
plot_spp <- function(x) {
  x %>% 
    filter(species %in% spp) %>% 
  ggplot(aes(census, ba)) +
  geom_col(aes(fill = species), position = "dodge")
}
```

Figure from data by Suzanne.

```{r}
lao %>% 
  mutate(census = plotcensusnumber, ba = ba_over_50) %>% 
  plot_spp() +
  labs(title = "Suzanne's data")
```

Compare with figure from data calculated with `basal_area_byyr()` after fixing bug reported in issue 71 (https://github.com/forestgeo/fgeo.abundance/issues/71). Now it looks very similar to Suzanne's data.

```{r}
mao %>% 
  filter(species %in% spp) %>% 
  gather("years", "ba", `1982`:`2015`) %>% 
  mutate(census = as.integer(as.factor(years))) %>% 
  ggplot(aes(census, ba)) +
  geom_col(aes(fill = species), position = "dodge") +
  labs(title = "Mauro's data")
```

